Binning Strategies and Related Loss for Large Data

### 2015-04-24 Paper.Rnw
streamline the terminology about artificial stripes
restructured Discussion/Conclusion sections to remove redundancy

While bin widths can be set to any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot. If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.  These white-space gaps are undesirable because they create artificial discontinuity that interferes with interpretation of bivariate density. If bin dimensions are not integer multiples of the bivariate data resolution there is also a disruptive visual consequence of ``stripping'', where bins have systematically different numbers of possible data values that can cause artificial density stripes to appear in the binned scatterplot.

To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension. Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.  Note that random binning is effective at smoothing out the stripped patterns when non-integer multiples of the data resolution are selected.

@@

The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction. Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure. Ideally we would like to strike the optimal trade-off between bin size and spatial loss, but unfortunately the simulated data from the three bivariate distributions indicated that this relationship between them was nearly linear. Thus it is left to the plot designer to decide if they wish to highlight high level density structure with higher spatial loss or finer density structure with lower spatial loss.

For spatial binning, the random binning algorithm displayed strong advantage of avoiding the visual ``stripes'' problem that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution. However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid the stripe illusion if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.  It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.

If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades.  This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. The standard and quantile algorithms for frequency binning have very similar loss properties for this range, so the decision becomes one of how we wish the shade of a tile to be interpreted. Standard log frequency binning is an alternative if the bin counts are heavily right skewed and we wish to find distributional structure in bins with low density. We also discovered that if we elect to use quantile binning, the log transformation will have no effect on the binned scatterplot shading due to the identical quantile-based mapping to ordinal frequency groupings for all monotone transformations on bin counts.

These recommendations for binned scatterplot construction and loss properties will be demonstrated in the following examples using real bivariate data. First we will further explore the baseball pitching data. Additionally we will investigate the relationship between scheduled and actual departure times for American commercial airline data.

@@

\ktm{We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features. In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value. }

\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading. It is recommended to use less than ten separate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.  }

>>>>

While bin widths can be set to any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot. If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.  These white-space gaps are undesirable because they create visual discontinuity that interferes with interpretation of bivariate density. There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial stripping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.

To demonstrate the importance of properly selecting binwidths consider the coarse version of the simulated uniform data which is recorded to a resolution of two units in each dimension. Figure~\ref{fig:UniformStripes} displays the binned scatterplots under several scenarios. Under standard binning we see the white-space gaps with one-by-one unit bins, vertical and horizontal artificial stripes with five-by-five unit bins and the appropriate view of an evenly spread density with four-by-four unit bins.  Note that random binning is effective at smoothing out the artificial stripping patterns when non-integer multiples of the data resolution are selected.

@@

The exploration of loss of spatial and frequency information in the process of creating a binned scatterplot yielded a number of important properties and practical recommendations for their construction. These recommendations and loss properties will now be demonstrated through two examples using real bivariate data. First we further explore the baseball pitching data. Additionally we will investigate the relationship between scheduled and actual departure times for American commercial airline data.

@@

\ktm{We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms. The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occurred when data recorded to a coarse resolution was binned using a bin width that was a non-integer multiple of the resolution. However, standard binning algorithm is superior due to lower spatial loss, and data with coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) can avoid creating artificial stripes if a bin dimensions are integer multiples of $\alpha_x$ and $\alpha_y$.  It was also shown through simulation that for coarse data, a reasonable default for the binning origin is offset to ($\alpha_x/2$,$\alpha_y/2$) because it resulted in minimal or near minimal spatial loss for symmetric data and preformed well even for skewed data.  }

\ktm{Spatially binning with smaller bin dimensions will lead to lower spatial losses; however, finer binning requires more processing time and does not highlight large scale density structure. It is left to the plot designer to decide how much spatial information they are willing to sacrific in order to simplify the display of density structure; we do however suggest to err on the side of smaller bins which allow for visualization of finer density structure with lower spatial loss. }

\ktm{If we elect to use frequency binning to aid the perceptual mapping tile shades to the plot key for frequency, it is recommended to use between four and seven distinct shades. This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions. }
